\documentclass[12pt,a4paper]{article}

% Encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Layout
\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\usepackage{calc}

% Links and references
\usepackage[colorlinks=true,linkcolor=blue!70!black,urlcolor=blue!70!black,citecolor=blue!70!black,breaklinks=true]{hyperref}
\usepackage{bookmark}

% Code listings with Go syntax highlighting
\usepackage{listings}
\usepackage{xcolor}

\definecolor{goKeyword}{RGB}{0,0,180}
\definecolor{goComment}{RGB}{106,115,125}
\definecolor{goString}{RGB}{163,21,21}
\definecolor{goType}{RGB}{38,127,153}
\definecolor{goBackground}{RGB}{248,248,248}
\definecolor{goLineNum}{RGB}{160,160,160}

\lstdefinelanguage{Gno}{
  language=Go,
  morekeywords={realm,cross,cur,attach,safely,revive},
  morekeywords=[2]{string,int,bool,any,error,byte,rune,uint64},
  morekeywords=[3]{true,false,nil,iota},
  sensitive=true,
}

\lstset{
  language=Gno,
  basicstyle=\ttfamily\footnotesize\linespread{0.9}\selectfont,
  keywordstyle=\color{goKeyword}\bfseries,
  keywordstyle=[2]\color{goType},
  keywordstyle=[3]\color{goKeyword},
  commentstyle=\color{goComment},
  stringstyle=\color{goString},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{goBackground},
  showstringspaces=false,
  tabsize=4,
  numbers=left,
  numberstyle=\tiny\color{goLineNum},
  numbersep=8pt,
  xleftmargin=2em,
  framexleftmargin=1.5em,
  captionpos=b,
  aboveskip=0.8em,
  belowskip=0.4em,
  rulecolor=\color{gray!40},
  columns=fullflexible,
  keepspaces=true,
}

% Tables and lists
\usepackage{longtable}
\usepackage{array}
\usepackage{enumitem}

% Graphics
\usepackage{graphicx}

% Allow deep nesting of itemize (up to 9 levels)
\setlistdepth{9}
\renewlist{itemize}{itemize}{9}
\setlist[itemize,1]{label=\textbullet}
\setlist[itemize,2]{label=--}
\setlist[itemize,3]{label=*}
\setlist[itemize,4]{label=\textperiodcentered}
\setlist[itemize,5]{label=$\diamond$}
\setlist[itemize,6]{label=$\circ$}
\setlist[itemize,7]{label=\textbullet}
\setlist[itemize,8]{label=--}
\setlist[itemize,9]{label=*}

% Pandoc compatibility
\providecommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\passthrough}[1]{#1}

\title{\Huge\textbf{Gno.land} \\ \Large Whitepaper}
\author{Jae Kwon}
\date{March 9, 2026}

\begin{document}

\maketitle
\tableofcontents
\newpage

\section{Executive Summary}\label{executive-summary}

Gno is the first general-purpose multi-user programming language. It
extends Go---a language used by millions of developers---with
language-level primitives for multi-user programming: explicit realm
boundaries, automatic object persistence, and a shared garbage-collected
heap across trust domains.

Every existing programming language assumes a single trusted user.
Interactions between users' programs require extra-language frameworks
(RPC, serialization, middleware) that add complexity and surface area
for exploits. Gno eliminates this: importing and calling another user's
code uses the same syntax and type system as importing a library.

\textbf{Key technical contributions:}

\begin{itemize}
\item
  \textbf{Interrealm specification.} A formal model for multi-user
  program interaction with explicit crossing semantics
  (\passthrough{\lstinline!cross!}/\passthrough{\lstinline!cur realm!}),
  readonly taints on external objects, and realm-transaction
  finalization. Every trust boundary is visible in source code and
  verifiable by the compiler.
\item
  \textbf{Automatic object persistence.} Ordinary Go structs persist
  automatically across transactions with Merkle-ization and garbage
  collection handled by the runtime. No databases, ORMs, or
  serialization code required.
\item
  \textbf{Shared object graph.} Objects in one realm can reference
  objects in another realm with language-level type safety. The garbage
  collector spans trust boundaries. No
  \passthrough{\lstinline!externref!}, no manual memory management, no
  custom serialization.
\item
  \textbf{Immutable packages as trust anchors.} Pure packages
  (\passthrough{\lstinline!/p/!}) are immutable on-chain, enabling two
  mutable realms to trustlessly cooperate through shared immutable
  code---a guarantee no other smart contract platform provides at the
  package-system level.
\end{itemize}

\textbf{Gno vs alternatives:} Solidity's implicit
\passthrough{\lstinline!msg.sender!} changes have caused billions in
exploits; Move requires a bespoke language with no shared heap;
WASM-based systems cannot persist individual objects without serializing
entire memory. Gno gives Go developers a familiar language with
multi-user safety built in.

Gno.land is a blockchain built on Tendermint2 that runs the GnoVM,
serving as the first multi-user language-based operating system. This
paper specifies the interrealm model, the Gno language extensions, and
the Gno.land blockchain architecture.

\newpage

\section{Interrealm Specification}\label{interrealm-specification}

\subsection{Introduction}\label{introduction}

All modern popular programming languages are designed for a single
programmer user. Programming languages support the importing of program
libraries natively for components of the single user's program, but this
does not hold true for interacting with components of another user's
(other) program. Gno is an extension of the Go language for multi-user
programming. Gno allows a massive number of programmers to iteratively
and interactively develop a single shared program such as Gno.land.

The added dimension of the program domain means the language should be
extended to best express the complexities of programming in the
inter-realm (inter-user) domain. In other words, Go is a restricted
subset of the Gno language in the single-user context. (In this analogy
client requests for Go web servers don't count as they run outside the
server program).

\subsubsection{Interrealm Programming
Context}\label{interrealm-programming-context}

Gno.land supports three types of packages: - \textbf{Realms
(\passthrough{\lstinline!/r/!})}: Stateful user applications (smart
contracts) that maintain persistent state between transactions -
\textbf{Pure Packages (\passthrough{\lstinline!/p/!})}: Stateless
libraries that provide reusable functionality - \textbf{Ephemeral
Packages (\passthrough{\lstinline!/e/!})}: Temporary code execution with
MsgRun which allows a custom main() function to be run instead of a
single function call as with MsgExec.

For an overview of the different package types in Gno
(\passthrough{\lstinline!/p/!}, \passthrough{\lstinline!/r/!}, and
\passthrough{\lstinline!/e/!}), see
\href{../builders/anatomy-of-a-gno-package.md}{Anatomy of a Gno
Package}.

Interrealm programming refers to the ability of one realm to call
functions in another realm. This can occur between: - Regular realms
(\passthrough{\lstinline!/r/!}) calling other regular realms via MsgExec
and MsgRun. - Ephemeral realms (\passthrough{\lstinline!/e/!}) calling
regular realms via MsgRun (like main.go)

The key concept is that code executing in one realm context can interact
with and call functions in other realms while leveraging the language
syntax rules of Go, enabling complex multi-user interactions while
maintaining clear boundaries and permissions.

\subsubsection{Realm-Context and
Realm-Storage-Context}\label{realm-context-and-realm-storage-context}

All logic in Gno executes under two contexts that together govern
identity and persistence:

\textbf{Realm-context} determines
\passthrough{\lstinline!runtime.CurrentRealm()!} and
\passthrough{\lstinline!runtime.PreviousRealm()!}. It controls identity
and agency: who is the current actor and who called them. The
realm-context has an associated Gno address from which native coins can
be sent and received. It changes only on explicit cross-calls
(\passthrough{\lstinline!fn(cross, ...)!}).

\textbf{Realm-storage-context} determines where new and modified objects
are persisted during realm-transaction finalization. It changes on
explicit cross-calls \emph{and} on implicit borrow-crosses (calling a
non-crossing method on a real object residing in a different realm). The
realm-storage-context is not directly accessible at runtime and has no
associated address.

After an explicit cross-call, both contexts refer to the same realm.
They diverge when calling a non-crossing method of a real object
residing in a different realm --- the realm-storage-context shifts to
the receiver's realm while the realm-context stays the same.

{\small
\begin{longtable}{|p{4.2cm}|c|c|c|c|}
\hline
\textbf{Call type} & \textbf{Realm-ctx?} & \textbf{Storage-ctx?} & \textbf{Boundary?} & \textbf{Finalizes?} \\
\hline
\endfirsthead
\hline
\textbf{Call type} & \textbf{Realm-ctx?} & \textbf{Storage-ctx?} & \textbf{Boundary?} & \textbf{Finalizes?} \\
\hline
\endhead
\lstinline!fn(cross,...)! to same realm & Yes* & No & Yes & Yes \\
\hline
\lstinline!fn(cross,...)! to diff.\ realm & Yes & Yes & Yes & Yes \\
\hline
\lstinline!fn(nil,...)! (non-crossing) & No & No & No & No \\
\hline
Non-crossing method, real recv.\ in same realm & No & No & No & No \\
\hline
Non-crossing method, real recv.\ in diff.\ realm & No & Yes & Yes & Yes \\
\hline
Non-crossing method, unreal recv. & No & No & No & No \\
\hline
Non-crossing function & No & No & No & No \\
\hline
\end{longtable}
}

* \passthrough{\lstinline!runtime.CurrentRealm()!} returns the same
realm, but \passthrough{\lstinline!runtime.PreviousRealm()!} shifts ---
what was current becomes previous. See \hyperref[realm-boundaries]{Realm
Boundaries} for definitions of boundary and finalization.

\subsubsection{Design Goals}\label{design-goals}

\textbf{Caveat: The interrealm specification does not secure
applications against arbitrary code execution. It is important for realm
logic (and even p package logic) to ensure that arbitrary (variable)
functions (and similarly arbitrary interface methods) are not provided
by malicious callers; such arbitrary functions and methods whether
crossing (or non-crossing) will inherit the previous realm (or both
current and previous realms) and could abuse these realm-contexts.} It
does not make sense for any realm user to cross-call an arbitrary
function or method as it loses agency while being marked as the
responsible caller by the callee's runtime previous realm. This problem
is worse when calling a non-crossing function or method. It can be
reasonable when such variable functions or interface values are
restricted in other ways such as by whitelisting by a DAO upon careful
inspection of every such variable function or interface value (both its
type declaration as well as its state).

P package code should behave the same even when copied verbatim in a
realm package; and likewise non-crossing code should behave the same
when copied verbatim from one realm to another. Otherwise there will be
lots of security related bugs from user error.

Realm crossing with respect to
\passthrough{\lstinline!runtime.CurrentRealm()!} and
\passthrough{\lstinline!runtime.PreviousRealm()!} must be explicit and
warrants type-checking; because a crossing-function of a realm should be
able to call another crossing-function of the same realm without
necessarily crossing (changing the realm-context). Sometimes the
previous realm and current realm must be the same realm, such as when a
realm consumes a service that it offers to external realms and users.

For Go developers familiar with blockchain VMs like the EVM: in
Solidity, calling another contract implicitly shifts
\passthrough{\lstinline!msg.sender!}, making it easy to introduce
reentrancy bugs or misattribute the caller. Gno's
\passthrough{\lstinline!cross!} keyword makes every realm-context
transition visible in source code and verifiable by the compiler,
eliminating this class of bugs by construction.

Where a real object resides should not matter too much, as it is often
difficult to predict. Thus the realm-context as returned by
\passthrough{\lstinline!runtime.PreviousRealm()!} and
\passthrough{\lstinline!runtime.CurrentRealm()!} should not change with
non-crossing method calls, and the realm-storage-context should be
determined for non-crossing methods only by the realm-storage of the
receiver. The realm-storage of a receiver should only matter for when
elements reside in external realm-storage and direct dot-selector or
index-expression access of sub-elements are desired of the
aforementioned element.

A method should be able to modify the receiver and associated objects of
the same realm-storage as that of the receiver.

A method should be able to create new objects that reside in the same
realm by association in order to maintain storage realm consistency and
encapsulation and reduce fragmentation.

It is difficult to migrate an object from one realm to another even when
its ref-count is 1; such an object may be deep with many descendants of
ref-count 1 and so performance is unpredictable.

Code declared in p packages (or declared in ``immutable'' realm
packages) can help different realms enforce contracts trustlessly, even
those that involve the caller's current realm. Otherwise two mutable
(upgradeable) realms cannot export trust unto the chain because
functions declared in those two realms can be upgraded. This is
analogous to Go interfaces but stronger: a Go interface guarantees
method signatures, while a Gno p-package guarantees \emph{behavior} ---
the implementation is immutable on-chain, so both parties can trust the
logic without trusting each other.

Both \passthrough{\lstinline!fn(cross, ...)!} and
\passthrough{\lstinline!func fn(cur realm, ...)\{...\}!} may become
special syntax in future Gno versions.

\begin{lstlisting}[language=Go]
// realm /r/alice/alice
package alice

var object any

func SetObject(cur realm, obj any) {
    object = obj
}
\end{lstlisting}

\begin{lstlisting}[language=Go]
// package /p/bob/types
package types

type UserProfile struct {
    Name string
    ...
}
\end{lstlisting}

\begin{lstlisting}[language=Go]
// realm /r/bob/bob
package bob

import "gno.land/r/alice/alice" // import external realm package
import "gno.land/p/bob/types"   // import external library package

func Register(cur realm, name string) {
    prof := types.UserProfile{Name: name}
    alice.SetObject(cross, prof)
}
\end{lstlisting}

The Gno language is extended to support a
\passthrough{\lstinline!context.Context!}-like argument to denote the
current realm-context of a Gno function. This allows a user realm
function to call itself safely as if it were being called by an external
user, and helps avoid a class of security issues that would otherwise
exist.

\begin{lstlisting}[language=Go]
// realm /r/alice/mail

func SendMail(cur realm, text string) {
    if text == "" {
        // runtime.PreviousRealm() is preserved for recursive call.
        SendMail(nil, "<empty>")
    }
    caller := runtime.PreviousRealm()
    if inBlacklist(caller) {
        // runtime.PreviousRealm() becomes self; message from self to self.
        SendMail(cross, fmt.Sprintf("blacklisted caller %v blocked", caller))
    } else {
        // sendMailPrivate not exposed to external callers.
        sendMailPrivate(text)
    }
}
\end{lstlisting}

\subsection{Object Model and Realm
Storage}\label{object-model-and-realm-storage}

Unlike other blockchain platforms where developers must manually
serialize state into key-value stores, Gno persists ordinary Go structs
and slices automatically. A Go developer can write familiar data
structures --- trees, linked lists, maps of structs --- and they survive
across transactions with no explicit marshaling. The runtime handles
Merkle-ization and garbage collection transparently.

Every object in Gno is persisted on disk with additional metadata
including the object ID and an optional OwnerID (if persisted with a
ref-count of exactly 1). The object ID is only set at the end of a
realm-transaction during realm-transaction finalization (more on that
later). A GnoVM transaction is composed of one or many scoped (stacked)
realm-transactions.

\begin{lstlisting}[language=Go]
type ObjectInfo struct {
    ID       ObjectID  // set if real.
    Hash     ValueHash `json:",omitempty"` // zero if dirty.
    OwnerID  ObjectID  `json:",omitempty"` // parent in the ownership tree.
    ModTime  uint64    // time last updated.
    RefCount int       // for persistence. deleted/gc'd if 0.

    // Object has multiple references (refcount > 1) and is persisted separately
    IsEscaped bool `json:",omitempty"` // hash in iavl.
    ...
}
\end{lstlisting}

When an object is persisted during realm-transaction finalization the
object becomes ``real'' (as in it is really persisted in the virtual
machine state) and is said to ``reside'' in the realm; and otherwise is
considered ``unreal''. New objects instantiated during a transaction are
always unreal; and during finalization such objects are either discarded
(transaction-level garbage collected) or become persisted and real.

Unreal (new) objects that become referenced by a real (persisted) object
at runtime will get their OwnerID set to the parent object's storage
realm, but will not yet have its object ID set before realm-transaction
finalization. Subsequent references at runtime of such an unreal object
by real objects residing in other realms do not override the OwnerID
initially set, so during realm-transaction finalization it ends up
residing in the first realm it became associated with (referenced from).
Unreal objects that become persisted but were never directly referenced
by any real object during runtime will only get their OwnerID set to the
realm of the first real ancestor.

Real objects with ref-count of 1 have their hash included in the sole
parent object's serialized byte form, thus an object tree of only
ref-count 1 descendants are Merkle-hashed completely.

When a real or unreal object ends up with a ref-count of 2 or greater
during realm-transaction finalization its OwnerID is set to zero and the
object is considered to have ``escaped''. When such a real object is
persisted with ref-count of 2 or greater it is forever considered
escaped even if its ref-count in later transactions is reduced to 1.
Escaped real objects do not have their hash included in the parent
objects' serialized byte form but instead are Merkle-ized separately in
an iavl tree of escaped object hashes (keyed by the escaped object's ID)
for each realm package. (This is implemented as a stub but not yet
implemented for the initial release of Gno.land.)

\textbf{A real object can only be directly mutated through dot-selectors
and index-expressions if the object resides in the same realm as the
current realm-storage-context. Unreal objects can always be directly
mutated if their elements are directly exposed.}

Exposed values accessed through dot-selectors and index-expressions from
external realm logic are tainted read-only. For the full rules see
\hyperref[readonly-taint-specification]{Readonly Taint Specification}.

\subsection{Crossing-Functions and
Crossing-Methods}\label{crossing-functions-and-crossing-methods}

Realm crossing occurs when a crossing function (declared as
\passthrough{\lstinline!func fn(cur realm, ...)\{...\}!}) is called with
the Gno \passthrough{\lstinline!fn(cross, ...)!} syntax.

\begin{lstlisting}[language=Go]
package main
import "gno.land/r/alice/realm1"

func main() {
    bread := realm1.MakeBread(cross, "flour", "water")
}
\end{lstlisting}

(In Linux/Unix operating systems user processes can cross-call into the
kernel by calling special syscall functions, but user processes cannot
directly cross-call into other users' processes. This makes the GnoVM a
more complete multi-user operating system than traditional operating
systems.)

Besides explicit realm crossing via the
\passthrough{\lstinline!fn(cross, ...)!} Gno syntax, implicit realm
crossing occurs when calling a method of a receiver object stored in an
external realm. Implicitly crossing into (borrowing) a receiver object's
storage realm allows the method to directly modify the receiver as well
as all other objects directly reachable from the receiver stored in the
same realm as the receiver. Unlike explicit crosses, implicit crosses do
not shift or otherwise affect the current realm context;
\passthrough{\lstinline!runtime.CurrentRealm()!} does not change unless
a method is called like
\passthrough{\lstinline!receiver.Method(cross, args...)!}.

Realms hold objects in residence and they also have a Gno address to
send and receive coins from. Coins can only be spent from the current
realm context.

\subsubsection{Definition}\label{definition}

A crossing-function or crossing-method is that which is declared in a
realm and has as its first argument \passthrough{\lstinline!cur realm!}.
The \passthrough{\lstinline!cur realm!} argument must appear as the
first argument of a crossing-function or crossing-method's argument
parameter list. To prevent confusion it is illegal to use anywhere else,
and cannot be used in p packages.

The current realm-context and realm-storage-context changes when a
crossing-function or crossing-method is called with the
\passthrough{\lstinline!cross!} keyword in the first argument as in
\passthrough{\lstinline!fn(cross, ...)!}. Such a call is called a
``cross-call'' or ``crossing-call''.

\begin{lstlisting}[language=Go]
package main
import "gno.land/r/alice/extrealm"

func MyMakeBread(cur realm, ingredients ...any) { ... }

func main(cur realm) {
    MyMakeBread(cross, "flour", "water") // ok -- cross into self.
    extrealm.MakeBread(cross, "flour", "water") // ok -- cross into extrealm
}
\end{lstlisting}

When a crossing-function or crossing-method is called with
\passthrough{\lstinline!nil!} as the first argument instead of
\passthrough{\lstinline!cross!} it is called a non-crossing-call; and no
realm-context nor realm-storage-context change takes place.

\begin{lstlisting}[language=Go]
package main
import "gno.land/r/alice/extrealm"

func MyMakeBread(cur realm, ingredients ...any) { ... }

func main(cur realm) {
    MyMakeBread(nil, "flour", "water") // ok -- non-crossing.
    extrealm.MakeBread(nil, "flour", "water") // invalid -- external realm function
}
\end{lstlisting}

To prevent confusion a non-crossing-call of a crossing-function or
crossing-method declared in a realm different than that of the caller's
realm-context and realm-storage-context will result in either a
type-check error; or a runtime error if the crossing-function or
crossing-method is variable.

\subsubsection{Realm-Context Rules}\label{realm-context-rules}

All functions in Gno execute under a realm context as determined by the
call stack. Objects that reside in a realm can only be modified if the
realm context matches.

A function declared in p packages when called:

\begin{itemize}
\tightlist
\item
  inherits the last realm for package declared functions and closures.
\item
  inherits the last realm when a method is called on unreal receiver.
\item
  implicitly crosses to the receiver's resident realm when a method of
  the receiver is called. The receiver's realm is also called the
  ``borrow realm''.
\end{itemize}

A function declared in a realm package when called:

\begin{itemize}
\tightlist
\item
  explicitly crosses to the realm in which the function is declared if
  the function is declared as
  \passthrough{\lstinline!func fn(cur realm, ...)\{...\}!} (with
  \passthrough{\lstinline!cur realm!} as the first argument). The new
  realm is called the ``current realm''.
\item
  otherwise follows the same rules as for p packages.
\end{itemize}

\passthrough{\lstinline!runtime.CurrentRealm()!} returns the current
realm-context that was last cross-called to.
\passthrough{\lstinline!runtime.PreviousRealm()!} returns the
realm-context cross-called to before the last cross-call. All
cross-calls are explicit with the \passthrough{\lstinline!cross!}
keyword, as well as non-crossing-calls of crossing-functions and
crossing-methods with \passthrough{\lstinline!nil!} instead of
\passthrough{\lstinline!cross!}.

A crossing function declared in the same realm package as the callee may
be called like \passthrough{\lstinline!fn(cross, ...)!} or
\passthrough{\lstinline!fn(cur, ...)!}. When called with
\passthrough{\lstinline!fn(cur, ...)!} there will be no realm crossing,
but when called like \passthrough{\lstinline!fn(cross, ...)!} there is
technically a realm crossing and the current realm and previous realm
returned are the same.

The current realm and previous realm do not depend on any implicit
crossing to the receiver's borrowed/storage realm even if the borrowed
realm is the last realm of the call stack. In other words
\passthrough{\lstinline!runtime.CurrentRealm()!} may differ from the
Machine's active storage realm (internally
\passthrough{\lstinline!m.Realm!}, i.e.~the borrow realm) when a method
is called on a receiver residing in a foreign realm.

\subsubsection{Implicit Realm-Storage
Borrowing}\label{implicit-realm-storage-borrowing}

Besides (explicit) realm-context changes via the
\passthrough{\lstinline!fn(cross, ...)!} cross-call syntax, implicit
realm-storage-context changes occur when calling a non-crossing method
of a receiver object residing in different realm-storage. The
realm-storage-context ``borrows'' to the receiver's realm, which allows
the method to directly modify its receiver and any objects reachable
from it that reside in the same realm-storage. The realm-context does
not change (so \passthrough{\lstinline!runtime.CurrentRealm()!} and
\passthrough{\lstinline!runtime.PreviousRealm()!} are unaffected; the
agency of the caller remains the same). However, objects reachable from
the receiver but residing in a \emph{different} realm-storage cannot be
modified --- the \passthrough{\lstinline!DidUpdate()!} guard in the
runtime enforces that only objects belonging to the borrowed realm can
be mutated.

This borrowing behavior allows non-crossing methods of receivers to
behave the same whether declared in a realm package or p package: p
package code copied over to a realm package, or realm package code
copied over to another realm, have the exact same behavior. Crossing
methods of a realm package would still behave differently when copied
over to another realm, as crossing-methods always change the
realm-context and realm-storage-context to the declared realm.

\subsubsection{Crossing-Method
Semantics}\label{crossing-method-semantics}

When a method is a crossing-method called as
\passthrough{\lstinline!receiver.Method(cross, args...)!}, both the
realm-context and realm-storage-context change to that of the realm
package in which the type/method is declared (which is not necessarily
the same as where the receiver resides). Such a crossing method-call
cannot directly modify the real receiver if it happens to reside in an
external realm that differs from where the type and methods are
declared; but it can modify any unreal receiver or unreal reachable
objects. As mentioned previously a non-crossing-call of a
crossing-method will fail during type-checking or at runtime if the
receiver resides in an external realm-storage.

Calls of methods on receivers residing in realms different from the
current realm must \emph{not} be called like
\passthrough{\lstinline!fn(cross, ...)!} if the method is not a crossing
function itself, and vice versa. Or it could be said that implicit
crossing is not real realm crossing. (When you sign a document with
someone else's pen it is still your signature; signature:pen ::
current:borrowed)

A crossing method declared in a realm cannot modify the receiver if the
object resides in a different realm. However not all methods are
required to be crossing methods, and crossing methods may still read the
state of the receiver (and in general anything reachable is readable).

\begin{lstlisting}[language=Go]
// Type and methods declared in /r/alice/tokens
package tokens

type Token struct {
    Owner   string
    Balance int
}

// Non-crossing method: borrows receiver's storage realm.
func (t *Token) Debit(amount int) {
    t.Balance -= amount
}

// Crossing method: realm-context and realm-storage-context
// shift to /r/alice/tokens (where the type is declared).
func (t *Token) Transfer(cur realm, to string, amount int) {
    t.Balance -= amount // FAILS if receiver resides outside /r/alice/tokens
}
\end{lstlisting}

\begin{lstlisting}[language=Go]
// /r/bob/bob stores a Token
package bob

import "gno.land/r/alice/tokens"

var myToken *tokens.Token // persisted in /r/bob/bob

func Spend(_ realm) {
    // Non-crossing call: storage-context borrows to /r/bob/bob
    // (receiver's realm). Debit() CAN modify myToken.
    myToken.Debit(10) // ok: borrowed storage matches receiver

    // Crossing call: storage-context shifts to /r/alice/tokens
    // (where the type is declared), not /r/bob/bob (where myToken resides).
    // Transfer() CANNOT directly modify myToken.
    myToken.Transfer(cross, "g1...", 10) // fails: receiver in external realm
}
\end{lstlisting}

New unreal objects reachable from the borrowed realm (or current realm
if there was no method call that borrowed) become persisted in the
borrowed realm (or current realm) upon finalization of the foreign
object's method (or function). (When you put an unlabeled photo in
someone else's scrapbook the photo now belongs to the other person.) In
the future the \passthrough{\lstinline!attach()!} function will prevent
a new unreal object from being taken.

For how crossing rules apply to MsgCall, MsgRun, and package
initialization, see \hyperref[message-types-and-testing]{Message Types
and Testing}.

\subsection{Realm Boundaries}\label{realm-boundaries}

A realm boundary is defined as a change in realm in the call frame stack
from one realm to another, whether explicitly crossed with
\passthrough{\lstinline!fn(cross, ...)!} or implicitly borrow-crossed
into a different receiver's storage realm. A realm may cross into itself
with an explicit cross-call.

When a crossing-function or crossing-method is cross-called it shifts
the ``current'' runtime realm-context to the ``previous'' runtime
realm-context such that
\passthrough{\lstinline!runtime.PreviousRealm()!} returns what used to
be returned with \passthrough{\lstinline!runtime.CurrentRealm()!} before
the realm boundary. The current realm-storage-context is always set to
that of realm-context after cross-calling.

For which call types create boundaries, see the
\hyperref[realm-context-and-realm-storage-context]{summary table} above.
Every crossing-call creates a realm boundary even when there is no
resulting change in realm-context or realm-storage-context. A
non-crossing-call of a crossing-function or crossing-method
(\passthrough{\lstinline!fn(nil, ...)!}) never creates a realm boundary.

\subsection{Realm-Transaction
Finalization}\label{realm-transaction-finalization}

Realm-transaction finalization occurs when returning from a realm
boundary. When returning from a cross-call (with
\passthrough{\lstinline!cross!}) realm-transaction finalization will
occur even with no change of realm-context or realm-storage-context.
Realm-transaction finalization does NOT occur when returning from a
non-crossing-call of a method of an unreal receiver or a real receiver
that resides in the same realm-storage-context as that of the caller.

During realm-transaction finalization all new reachable objects are
assigned object IDs and stored in the current realm-storage-context; and
ref-count-zero objects deleted (full ``disk-persistent cycle GC'' will
come after launch); and any modified ref-counts and new Merkle hash root
computed.

\subsection{Readonly Taint
Specification}\label{readonly-taint-specification}

Go's language rules for value access through dot-selectors \&
index-expressions are the same within the same realm, but exposed values
through dot-selector \& index-expressions are tainted read-only when
performed by an external realm.

The readonly taint prevents the direct modification of real objects by
any logic, even from logic declared in the same realm as that of the
object's storage-realm.

A realm cannot directly modify another realm's objects without calling a
function that gives permission for the modification to occur.

For example \passthrough{\lstinline!externalrealm.Foo!} is a
dot-selector expression on an external object (package) so the value is
tainted with the \passthrough{\lstinline!N\_Readonly!} attribute.

The same is true for \passthrough{\lstinline!externalobject.FieldA!}
where \passthrough{\lstinline!externalobject!} resides in an external
realm.

The same is true for \passthrough{\lstinline!externalobject[0]!}: direct
index expressions also taint the resulting value with the
\passthrough{\lstinline!N\_Readonly!} attribute.

The same is true for
\passthrough{\lstinline!externalobject.FieldA.FieldB[0]!}: the readonly
taint persists for any subsequent direct access, so even if FieldA or
FieldB resided in the caller's own realm-context or realm-storage the
result is tainted readonly.

A Gno package's global variables even when exposed
(e.g.~\passthrough{\lstinline!package realm1; var MyGlobal int = 1!})
are safe from external manipulation
(e.g.~\passthrough{\lstinline!import "xxx/realm1"; realm1.MyGlobal = 2!})
by the readonly taint when accessed directly by dot-selector or
index-expression from external realm logic; and also by a separate
\passthrough{\lstinline!DidUpdate()!} guard when accessed by other means
such as by return value of a function and the return value is real and
external.

A function or method's arguments and return values retain and pass
through any readonly taint from caller to callee. Even if a realm's
function (or method) returns an untainted real object, the runtime guard
in \passthrough{\lstinline!DidUpdate()!} prevents it from being modified
by an external realm-storage-context.

For a realm (user) to manipulate an untainted object residing in an
external realm, a function (or method) can be declared in the external
realm which references and modifies the aforementioned untainted object
directly (by a name declared outside of the scope of said function or
method). Or, the function can take in as argument an untainted real
object returned by another function.

Besides protecting against writing by direct access, the readonly taint
also helps prevent a class of security issue where a realm may be
tricked into modifying something that it otherwise would not want to
modify. Since the readonly taint prohibits mutations even from logic
declared in the same realm, it protects realms against mutating its own
object that it doesn't intend to: such as when a realm's real object is
passed as an argument to a mutator function where the object happens to
match the type of the argument.

Objects returned from functions or methods are not readonly tainted. So
if
\passthrough{\lstinline!func (eo object) GetA() any \{ return eo.FieldA \}!}
then \passthrough{\lstinline!externalobject.GetA()!} returns an object
that is not tainted assuming eo.FieldA was not otherwise tainted. While
the parent object \passthrough{\lstinline!eo!} is still protected from
direct modification by external realm logic, the returned object from
\passthrough{\lstinline!GetA()!} can be passed as an argument to logic
declared in the residing realm of \passthrough{\lstinline!eo.FieldA!}
for direct mutation.

Whether or not an object is readonly tainted it can always be mutated by
a method declared on the receiver.

\begin{lstlisting}[language=Go]
// /r/alice

var blacklist []string

func GetBlacklist() []string {
    return blacklist
}

func FilterList(cur realm, testlist []string) { // blanks out blacklist items from testlist
    for i, item := range testlist {
        if contains(blacklist, item) {
            testlist[i] = ""
        }
    }
}
\end{lstlisting}

This is a toy example, but you can see that the intent of
\passthrough{\lstinline!FilterList()!} is to modify an externally
provided slice; yet if you call
\passthrough{\lstinline!alice.FilterList(cross, alice.GetBlacklist())!}
you can trick alice into modifying its own blacklist--the result is that
alice.BlackList becomes full of blank values.

With the readonly taint \passthrough{\lstinline!var Blacklist []string!}
solves the problem for you; that is, /r/bob cannot successfully call
\passthrough{\lstinline!alice.FilterList(cross, alice.Blacklist)!}
because \passthrough{\lstinline!alice.Blacklist!} is readonly tainted
for bob.

The problem remains if alice implements
\passthrough{\lstinline!func GetBlacklist() []string \{ return Blacklist \}!}
since then /r/bob can call
\passthrough{\lstinline!alice.FilterList(cross, alice.GetBlacklist())!}
and the argument is not readonly tainted.

Future versions of Gno may also expose a new modifier keyword
\passthrough{\lstinline!readonly!} to allow for return values of
functions to be tainted as readonly. Then with
\passthrough{\lstinline!func GetBlacklist() readonly []string!} the
return value would be readonly tainted for both bob and alice.

\subsection{\texorpdfstring{\texttt{panic()} and
\texttt{revive(fn)}}{panic() and revive(fn)}}\label{panic-and-revivefn}

\passthrough{\lstinline!panic()!} behaves the same within the same realm
boundary, but when a panic crosses a realm boundary (as defined in
\hyperref[realm-boundaries]{Realm Boundaries}) the Machine aborts the
program. This is because in a multi-user environment it isn't safe to
let the caller recover from realm panics that often leave the state in
an invalid state.

This would be sufficient, but we also want to write our tests to be able
to detect such aborts and make assertions. For this reason Gno provides
the \passthrough{\lstinline!revive(fn)!} builtin.

\begin{lstlisting}[language=Go]
abort := revive(func() {
    fn := func(_ realm) {
        panic("cross-realm panic")
    }
    fn(cross)
})
abort == "cross-realm panic"
\end{lstlisting}

\passthrough{\lstinline!revive(fn)!} will execute `fn' and return the
exception that crossed a realm boundary during finalization.

This is only enabled in testing mode (for now), behavior is only
partially implemented. In the future
\passthrough{\lstinline!revive(fn)!} will be available for non-testing
code, and the behavior will change such that
\passthrough{\lstinline!fn()!} is run in transactional (cache-wrapped)
memory context and any mutations discarded if and only if there was an
abort.

TL;DR: \passthrough{\lstinline!revive(fn)!} is Gno's builtin for STM
(software transactional memory).

\subsection{\texorpdfstring{\texttt{attach()}}{attach()}}\label{attach}

In future releases of Gno the \passthrough{\lstinline!attach()!}
function can be used to associate unreal objects to the current
realm-storage-context before being passed into a function declared in an
external realm package, or into a method of a real receiver residing in
an external realm-context.

\subsection{\texorpdfstring{\texttt{safely(cb\ func())}}{safely(cb func())}}\label{safelycb-func}

In future releases of Gno the
\passthrough{\lstinline!safely(cb func())!} function may be used to
clear the current and previous realm-context as well as any
realm-storage-context such that no matter what
\passthrough{\lstinline!cb func()!} does the caller does not yield
agency to the callee.

For now this can be simulated by implementing an (immutable
non-upgradeable) realm crossing-function that cross-calls into itself
once more before calling the callback function.

\subsection{Guidelines}\label{guidelines}

P package code cannot contain crossing functions. P package code also
cannot import R realm packages. But code can call named crossing
functions e.g. those passed in as parameters.

You must declare a public realm function to be a crossing function if it
is intended to be called by end users, because users cannot MsgCall
non-crossing functions (for safety/consistency) or p package functions
(there's no point).

Utility functions that are a common sequence of non-crossing logic can
be offered in realm packages as non-crossing functions. These can also
import and use other realm utility non-crossing functions; whereas p
packages cannot import realm packages at all. And convenience/utility
functions that are being staged before publishing as permanent p code
should also reside in upgradeable realms.

Generally you want your methods to be non-crossing. Because they should
work for everyone. They are functions that are pre-bound to an object,
and that object is like a quasi-realm in itself, that could possibly
reside and migrate to other realms. This is consistent with any p code
copied over to r realms; none of those methods would be crossing, and
behavior would be the same; stored in any realm, mostly non-crossing
methods that anyone can call. Why is a quasi-realm self-encapsulated
object in need to modify the realm in which it is declared, by crossing?
That's intrusive, but sometimes desired.

You can always cross-call a method from a non-crossing method if you
need it.

\subsection{Message Types and Testing}\label{message-types-and-testing}

\subsubsection{MsgCall}\label{msgcall}

MsgCall may only call crossing functions. This is to prevent potential
confusion for non-sophisticated users. Non-crossing calls of
non-crossing functions of other realms is still possible with MsgRun.

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/r/test/test

func Public(_ realm) {

    // Returns (
    //     addr:<origin_caller>,
    //     pkgpath:""
    // ) == testing.NewUserRealm(origin_caller)
    runtime.PreviousRealm()

    // Returns (
    //     addr:chain.PackageAddress("gno.land/r/test/test"),
    //     pkgpath:"gno.land/r/test/test"
    // ) == testing.NewCodeRealm("gno.land/r/test/test")
    runtime.CurrentRealm()

    // Call a crossing function of same realm with crossing
    AnotherPublic(cross)

    // Call a crossing function of same realm without crossing
    AnotherPublic(cur)
}

func AnotherPublic(_ realm) {
    ...
}
\end{lstlisting}

\subsubsection{MsgRun}\label{msgrun}

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/e/g1user/run

import "gno.land/r/realmA"

func main() {
    // Before main() is called there is an implicit
    // crossing from UserRealm(g1user) to
    // CodeRealm(gno.land/e/g1user/run).

    // Returns (
    //     addr:g1user,
    //     pkgpath:""
    // ) == testing.NewUserRealm(g1user)
    runtime.PreviousRealm()

    // Returns (
    //     addr:g1user,
    //     pkgpath:"gno.land/e/g1user/run"
    // ) == testing.NewCodeRealm("gno.land/e/g1user/run")
    runtime.CurrentRealm()

    realmA.PublicNoncrossing()
    realmA.PublicCrossing(cross)
}
\end{lstlisting}

Realm addresses are derived from package paths via
\passthrough{\lstinline!chain.PackageAddress()!} (defined in
\passthrough{\lstinline!gnovm/stdlibs/chain/address.gno!}):

\begin{itemize}
\tightlist
\item
  \passthrough{\lstinline!chain.PackageAddress("gno.land/r/name123/realm")!}
  --- bech32 from hash(path)
\item
  \passthrough{\lstinline!chain.PackageAddress("gno.land/e/g1user/run")!}
  --- bech32 substring ``g1user''
\end{itemize}

Therefore in the MsgRun file's \passthrough{\lstinline!init()!} function
the previous realm and current realm have different pkgpaths (the origin
caller always has empty pkgpath) but the address is the same.

\subsubsection{MsgAddPackage}\label{msgaddpackage}

A realm package's initialization (including
\passthrough{\lstinline!init()!} calls) executes with current
realm-context of itself.
\passthrough{\lstinline!runtime.PreviousRealm()!} refers to the package
deployer both in global var decls and inside
\passthrough{\lstinline!init()!} functions. After that the package
deployer is no longer provided, so packages need to remember the
deployer in the initialization phase if needed.

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/r/test/test

func init() {
    // Returns (
    //     addr:<origin_deployer>,
    //     pkgpath:""
    // ) == testing.NewUserRealm(origin_deployer)
    // Inside init() and global var decls
    // are the only time runtime.PreviousRealm()
    // returns the deployer of the package.
    // Save it here or lose it forever.
    runtime.PreviousRealm()

    // Returns (
    //     addr:chain.PackageAddress("gno.land/r/test/test"),
    //     pkgpath:"gno.land/r/test/test"
    // ) == testing.NewCodeRealm("gno.land/r/test/test")
    runtime.CurrentRealm()
}

// Same as in init().
var _ = runtime.PreviousRealm()
\end{lstlisting}

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/e/g1user/run

func init() {
    // Returns (
    //     addr:g1user,
    //     pkgpath:""
    // ) == testing.NewUserRealm(g1user)
    runtime.PreviousRealm()

    // Returns (
    //     addr:g1user,
    //     pkgpath:"gno.land/e/g1user/run"
    // ) == testing.NewCodeRealm("gno.land/e/g1user/run")
    runtime.CurrentRealm()
}
\end{lstlisting}

The same applies for pure package (\passthrough{\lstinline!/p/!})
initialization. During initialization and tests,
\passthrough{\lstinline!runtime.CurrentRealm()!} can return a package
path that starts with ``/p/''. This is because the package is
technically still mutable during its initialization phase. After
initialization, pure packages become immutable and cannot maintain
state.

\subsubsection{Testing overrides with
stdlibs/testing}\label{testing-overrides-with-stdlibstesting}

The
\passthrough{\lstinline!gnovm/tests/stdlibs/testing/context\_testing.gno!}
file provides functions for overriding frame details from Gno test code.

\passthrough{\lstinline!testing.SetRealm(testing.NewUserRealm("g1user"))!}
is identical to
\passthrough{\lstinline!testing.SetOriginCaller("g1user")!}. Both will
override the Gno frame to make it appear as if the current frame is the
end user signing with a hardware signer. Both will also set
\passthrough{\lstinline!ExecContext.OriginCaller!} to that user. One of
these will become deprecated.

\paragraph{\texorpdfstring{Gno test cases with \texttt{\_test.gno} like
\texttt{TestFoo(t\ *testing.T)}}{Gno test cases with \_test.gno like TestFoo(t *testing.T)}}\label{gno-test-cases-with-_test.gno-like-testfoot-testing.t}

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/r/user/myrealm
package myrealm

import (
    "chain/runtime"
    "testing"
)

func TestFoo(t *testing.T) {
    // At first OriginCaller is not set.

    // Override the OriginCaller.
    testing.SetRealm(testing.NewUserRealm("g1user"))

    // Identical behavior:
    testing.SetOriginCaller("g1user")

    // This panics now: seeking beyond the overridden origin frame:
    // runtime.PreviousRealm()

    // Simulate g1user cross-calling Public().
    // Produce a new frame to override
    func() {
        testing.SetRealm(testing.NewCodeRealm("gno.land/r/user/myrealm"))

        runtime.PreviousRealm() // "g1user", ""
        runtime.CurrentRealm()  // bech32(hash("gno.land/r/user/myrealm")), "gno.land/r/user/myrealm"

        Public(...) // already in "gno.land/r/user/myrealm"
    }()

    // The following is identical to the above,
    // but not possible in p packages which
    // cannot import realms.
    Public(cross, ...)
}
\end{lstlisting}

\paragraph{\texorpdfstring{Gno filetest cases with
\texttt{\_filetest.gno}}{Gno filetest cases with \_filetest.gno}}\label{gno-filetest-cases-with-_filetest.gno}

\begin{lstlisting}[language=Go]
// PKGPATH: gno.land/r/test/test
package test

import (
    "chain/runtime"
    "testing"

    "gno.land/r/user/myrealm"
)

func init() {
    // XXX Frame not found, there is no deployer for filetests.
    runtime.PreviousRealm()

    // Returns (
    //     addr:chain.PackageAddress("gno.land/r/test/test")
    //     pkgpath:"gno.land/r/test/test"
    // ) == testing.NewCodeRealm("gno.land/r/test/test")
    runtime.CurrentRealm()
}

func main() {
    // There is assumed to be in "frame -1"
    // a crossing from UserRealm(g1user) to
    // CodeRealm("gno.land/r/test/test") before
    // main() is called, so crossing() here
    // is redundant.

    // Returns (
    //     addr:g1user,
    //     pkgpath:""
    // ) == testing.NewUserRealm(g1user)
    runtime.PreviousRealm()

    // Returns (
    //     addr:g1user,
    //     pkgpath:"gno.land/r/test/test"
    // ) == testing.NewCodeRealm("gno.land/r/test/test")
    runtime.CurrentRealm()

    // gno.land/r/test/test cross-calling
    // gno.land/r/user/myrealm:
    myrealm.Public(cross, ...)
}

// Output:
// XXX
\end{lstlisting}

\subsection{Implementation}\label{implementation}

Implementation for \passthrough{\lstinline!runtime.CurrentRealm()!} and
\passthrough{\lstinline!runtime.PreviousRealm()!} are defined in
\passthrough{\lstinline!gnovm/stdlibs/chain/runtime/native.gno!} and
related files in the directory, while overrides for testing are defined
in
\passthrough{\lstinline!gnovm/tests/stdlibs/testing/context\_testing.gno!}.
All stdlibs functions are available unless overridden by the latter.

\subsection{Proposed Changes}\label{proposed-changes}

\passthrough{\lstinline!testing.SetOriginCaller()!} may be deprecated in
favor of
\passthrough{\lstinline!testing.SetRealm(testing.NewOriginRealm(user))!}.

\passthrough{\lstinline!testing.NewCodeRealm(path)!} may be renamed to
\passthrough{\lstinline!testing.NewPackageRealm(path)!}.

\newpage

\section{Gno Language}\label{gno-language}

\emph{We could not have arrived at the final design of the Gno language
without the help of the many developers who gave much time of their
lives to contribute to this project and the design of the language;
specifically the NewTendermint Gno team and Gno community, and
especially Manfred, Morgan, Maxwell, Milos, Guilhem, Ray, and Omar.}

\textbf{Gno is the first general purpose \emph{multi-user} programming
language.}

By ``multi-user'' we mean a language where (1) multiple users' code and
data coexist in a shared persistent heap, (2) trust boundaries between
users are enforced by the language itself--not by the operating system,
a hypervisor, or ad-hoc middleware--and (3) inter-user interoperability
uses the same syntax and type system as intra-user code. Prior systems
have achieved subsets of these properties: MUMPS (1966) shares
persistent global state across users but has no language-level trust
boundaries; MOO (1990) and LPC (1989) provide multi-user object
ownership in virtual worlds but are domain-specific MUD languages, not
general purpose; Erlang (1986) excels at concurrency but all processes
are equally trusted with no per-user security model; Solidity (2015) and
Move (2019) handle multi-user smart contracts but lack transparent
inter-module interoperability with a shared heap (as detailed
\hyperref[gno-vs-existing-smart-contract-platforms]{below}). Gno is the
first language to combine all three properties in a general purpose
language based on an existing mainstream language (Go).

Gno is a minimal extension of the Go language for multi-user
programming. Gno allows a massive number of programmers to iteratively
and interactively develop a single shared program such as Gno.land
(making it ideal for interoperable smart contract programming). In other
words, Go is a restricted subset of the Gno language in the single-user
context.

All of our programming languages to date are designed for a single
programmer user. All programming languages make the same assumption that
there is only one user -- the programmer, or program executor user.
Whether written in C, C++, Python, Java, Javascript, or Go, it is
assumed that all of the dependencies of the program are trusted. If
there is a vulnerability in any of the dependencies there is a
vulnerability in the program; it is the job of the programmer or
program/product manager to ensure that the overall program is free of
exploits.

When interacting with programs owned by another user (or process)
various techniques are used such as via IPC APIs often generated by
tools like Protobuf/GRPC; but such tools add extra complexity,
additional surface area for exploits, additional compute complexity, and
do not benefit directly from the language's native rules and
type-checker--especially for inter-process passing of in-memory object
references. In Gno, interacting with another realm's function is just
like interacting with any dependency (import and call).

Automatic persistence of memory objects makes databases, ORMs, and
serialization generally unnecessary. GnoVM at the end of realm
transaction boundaries will ``finalize'' by scanning all new, modified,
and deleted objects and will automatically persist what needs to be
saved along with meta information like the Merkle hash, ref count,
associated realm and unique number of each object. By making inter-user
interoperability native to the language and also by removing the
distinction between volatile and persistent memory, Gno eliminates much
of the accidental complexity that plagues traditional multi-service and
smart-contract development.

The GnoVM is implemented purely in Go and is a Gno AST interpreter. It
is implemented to serve as a reference implementation suitable for
production and tinkering. The GnoVM opcodes are expected to change over
time to balance simplicity, efficiency, and transpilation
targets--perhaps even to hardware. Just as Ethereum was originally
implemented in Python we expect future software to be optimized, whether
implemented in other languages, or possibly transpiled from the existing
GnoVM implementation.

For the latest Gno specs including the inter-realm spec refer to the
\href{https://github.com/gnolang/gno/tree/master/docs/resources}{Gno
docs/resources directory}.

\subsection{Gno vs Existing Smart Contract
Platforms}\label{gno-vs-existing-smart-contract-platforms}

Smart contract platforms like Ethereum allow for many users to upload
their application and call other user application logic functions, but
Solidity is not a general purpose programming language and has severe
limitations that make it suboptimal for solving the task at hand.

First, \textbf{Solidity and other existing smart-contract
languages/platforms do not support transparent inter-module (inter-user)
interoperability with the same language rules as for intra-module
dependencies.} That is, an application developer for a smart-contract
cannot simply import and call/use another user application's functions
and types as if they were library dependencies of the same application.
Generally interoperability between different modules are implemented
with extra-language frameworks and libraries on top of an incomplete or
primitive message-passing agent architecture; such interop function
calls generally do not share the same call-stack nor memory space.
Solidity's implicit \passthrough{\lstinline!msg.sender!} changes on
external calls have caused billions of dollars in exploits (reentrancy,
confused-deputy attacks). In Gno the programmer writes
\passthrough{\lstinline!cross!} at each call site that changes the
realm-context, making every trust boundary visible in the source code
and checkable by the compiler--a small syntactic cost for a huge safety
gain. Move (used by Sui and Aptos) improves on Solidity with a linear
type system that prevents reentrancy, but at the cost of a bespoke
language unfamiliar to most developers and without a shared persistent
object graph; inter-module calls in Move still pass values by copy or
move rather than by shared reference.

Second, \textbf{Solidity and other existing smart-contract
languages/platforms do not support the automatic persistence and
Merkle-ization of in-memory (heap) objects and often require custom
serialization logic.} Solidity does not support such a heap at all as
all memory for variables are predeclared in the function and as such is
not object-oriented and does not have a garbage collector or similar
memory-management primitives. WASM-based smart contract systems do not
support automatic persistence of objects without persisting the entire
memory state of the module. This requires a specialized virtual machine
such as the Gno VM which keeps track of every object created, modified,
and deleted. Solidity developers manage storage slots and mappings
manually; Move developers serialize structs into global storage with
explicit \passthrough{\lstinline!move\_to!} and
\passthrough{\lstinline!borrow\_global!} calls. Gno developers write
ordinary Go structs and they persist automatically with
Merkle-ization--a major reduction in complexity for the common case.

The automatic persistence of in-memory objects of the GnoVM is like a
memristor simulator. The advent of AI has created a new market for
memristor-based memory systems where the distinction between RAM
volatile memory and persistent disk storage is removed. Urbit is similar
but is not based on any general purpose programming language. With
memristor-based memory the GnoVM can be further simplified and the
performance of applications can be vastly improved without any changes
to the Gno language specification.

Third, \textbf{Solidity and other existing smart-contract
languages/platforms do not support a shared heap memory space for
objects to be referenced by external-user objects in a uniform manner by
language rules}. Alice cannot simply declare a structure object that
references the structure object persisted in Bob's application and trust
the garbage collector to retain Bob's object for as long as Alice's
object is retained.

Fourth, \textbf{the split between immutable p-packages
(\passthrough{\lstinline!gno.land/p/!}) and mutable r-realms
(\passthrough{\lstinline!gno.land/r/!}) solves a real trust problem that
no other smart contract platform addresses at the package-system level.}
Two mutable realms cannot trustlessly cooperate because either could
upgrade its code, but an immutable p-package can enforce a contract
between them. In Solidity, library contracts can be proxied and
upgraded, undermining any guarantees they were supposed to provide. In
Move, modules are also upgradeable by default. Gno bakes the
immutability guarantee into the package path convention itself.

\textbf{The above differentiating factors of the Gno language allows for
the most succinct expression of a single-user application or multi-user
application composed of independent modules without the extra complexity
from extra-language interop type-checking syntax or frameworks nor of
the extra complexity from any database, ORM, or serialization logic.}

Shared garbage-collection in a shared (multi-user) graph of object
references makes it possible for one's object representing (say) a
propositional statement or idea to be easily referenced by an
alternative statement or idea, or even be extended by reference with
additional commentary, metadata, or even a subreddit-like tree of
discussions. Without a shared garbage collector the task of ensuring
that references still hold over time without becoming dangling pointers
is left up to each inter-application interface at best, requiring custom
logic just to handle garbage collection. WebAssembly (WASM) externref
support in Go has limitations, particularly in how it handles external
memory references. Currently, Rust and Go do not natively support
externref types for function parameters or return values, making it
challenging to pass complex data between Wasm modules and their host
environments effectively.

\begin{quote}
Reference type (aka externref or anyref) is an opaque reference made
available to a WASM module by the host environment. Such references
cannot be forged in the WASM code and can be associated with arbitrary
host data, thus making them a good alternative to ad-hoc handles (e.g.,
numeric ones). References cannot be stored in WASM linear memory; they
are confined to the stack and tables with externref elements.

Rust does not support reference types natively; there is no way to
produce an import / export that has externref as an argument or a return
type. wasm-bindgen patches WASM if externrefs are enabled. This library
strives to accomplish the same goal for generic low-level WASM ABIs
(wasm-bindgen is specialized for browser hosts).

\textbf{externref use cases} Since externrefs are completely opaque from
the module perspective, the only way to use them is to send an externref
back to the host as an argument of an imported function. (Depending on
the function semantics, the call may or may not consume the externref
and may or may not modify the underlying data; this is not reflected by
the WASM function signature.) An externref cannot be dereferenced by the
module, thus, the module cannot directly access or modify the data
behind the reference. Indeed, the module cannot even be sure which kind
of data is being referenced. -
https://docs.rs/externref/latest/externref/
\end{quote}

Even if externref were fully implemented in future specs for Go (or
Rust) such that it could be used as an argument or return type across
modules (still not ideal for type-checking as it is not the underlying
type), this would limit what can be inter-module-referenced to that
which can be held in memory. The Gno Virtual Machine (GnoVM) allows for
inter-user-package (inter-realm) references across the entire persisted
disk store space, and does not require any additional language syntax
such as with the \passthrough{\lstinline!externref!} keyword, and
supports the normal course of type-checking already familiar to Go
developers.

\subsection{Interrealm Programming}\label{interrealm-programming}

Gno extends Go with explicit multi-user realm awareness. The key
concepts:

\begin{itemize}
\tightlist
\item
  \textbf{Realm-context}: determines
  \passthrough{\lstinline!runtime.CurrentRealm()!} and
  \passthrough{\lstinline!runtime.PreviousRealm()!}, controlling
  identity and agency.
\item
  \textbf{Realm-storage-context}: determines where new and modified
  objects are persisted during finalization.
\item
  \textbf{Crossing-functions}: declared with
  \passthrough{\lstinline!func fn(cur realm, ...)!} and called with
  \passthrough{\lstinline!fn(cross, ...)!} to explicitly change both
  contexts.
\item
  \textbf{Readonly taint}: values accessed via dot-selectors or
  index-expressions on external realm objects are tainted read-only,
  preventing direct mutation even by logic declared in the same realm.
  This protects against a class of exploits where a realm is tricked
  into mutating its own objects.
\item
  \textbf{Realm-transaction finalization}: at realm boundaries, new
  objects are assigned IDs and persisted, ref-count-zero objects are
  deleted, and Merkle hashes are recomputed.
\end{itemize}

\begin{lstlisting}[language=Go]
// realm /r/bob/bob
package bob

import "gno.land/r/alice/alice" // import external realm package
import "gno.land/p/bob/types"   // import external library package

func Register(cur realm, name string) {
    prof := types.UserProfile{Name: name}
    alice.SetObject(cross, prof) // explicit realm crossing
}
\end{lstlisting}

Non-crossing methods on receivers residing in external realms implicitly
change only the realm-storage-context (not the realm-context), allowing
methods to behave identically whether declared in a p package or realm
package. Panics that cross realm boundaries abort the program;
\passthrough{\lstinline!revive(fn)!} provides software transactional
memory semantics for testing (and eventually production use). Future
builtins include \passthrough{\lstinline!attach()!} for binding unreal
objects to a realm-storage-context, and
\passthrough{\lstinline!safely(cb)!} for clearing realm-contexts to
prevent yielding agency to callees.

For the full specification including the object model
(real/unreal/escaped objects), readonly taint rules, crossing semantics,
and design goals, see the
\href{./resources/gno-interrealm.md}{Interrealm Specification}.

\newpage

\section{Gno.land Blockchain}\label{gno.land-blockchain}

Tendermint solved proof-of-stake by innovating upon classical Byzantine
fault-tolerant consensus algorithms published by Dwork, Lynch, and
Stockmeyer in 1988 (originally funded by Darpa for missile defense
systems) for blockchains. It paved the way for the Cosmos Hub, the first
proof-of-stake IBC hub, and Cosmos the internet of blockchains. Also of
note, when Binance first launched they used the CosmosSDK and
Tendermint.

Gno.land builds upon Tendermint2 and aims to shift the paradigm of
programming languages in general: Gno is the first \emph{multi-user}
programming language, making it a superior smart contracting language as
compared to any existing solution. Thus \textbf{Gno.land is the first
multi-user language-based operating system}. Its ultimate goal is to be
the world's open knowledge base for the next millennium.

\subsection{Use Case: Open Programmable Knowledge
Base}\label{use-case-open-programmable-knowledge-base}

Go's simple embedded struct-centric design and the Gno VM's automatic
transactional persistence makes Gno.land not only great for
decentralized financial applications but also makes it uniquely well
suited and designed for permissionless innovation of information-based
applications such as social communication and coordination systems, or
the next Wikipedia or programmable knowledge-base systems. The latter
will be explored here.

Each of the thought statements in the
\hyperref[gno-land-for-mass-awakening]{introduction} can be represented
as a simple Go string, but as in Tractatus we want to allow each of
these thought- statements to be supported by any number of supporting
thought statements, so we need a struct declaration.

\begin{lstlisting}[language=Go]
type Thought struct {
    Statement    string
    Dependencies []*Thought
}
\end{lstlisting}

The above allows for a simple tree structure, but it would be better to
annotate each child node (thought statement) with the type of relation
to the parent node-- for example whether a child node represents an
example, a caveat, a corollary, or supporting evidence and so on.

\begin{lstlisting}[language=Go]
// Option "Denormalized Thought"
type Thought struct {
    Statement   string
    Examples    []*Thought
    Caveats     []*Thought
    Corollaries []*Thought
    Support     []*Thought
}
\end{lstlisting}

Better than a denormalized structure is one where the type of thought
statement association can be arbitrary or fixed depending on the
application.

\begin{lstlisting}[language=Go]
// Option "Normalized Thought"
type Thought struct {
    Text         string
    TypedSupport []*Thought
}

type ThoughtType string // examples, caveats, corollaries, support

type TypedThought struct {
    Type    ThoughtType
    Thought *Thought
}
\end{lstlisting}

\emph{Note on the usage of \passthrough{\lstinline![]\\*Thought!}
slices: in the current implementation of the GnoVM each slice can only
be used by first loading the entire underlying base array. This may be
optimized in the future, however for holding large sets of elements the
programmer should instead use a tree-structure such as the avl.Tree (or
an iavl.Tree).}

Now arises the question of whether counter-arguments should also be
referenced as a child node to the original thought parent node. If we
include counter-arguments in the graph of
\passthrough{\lstinline!*Thought!} objects itself there is the issue of
permissioning who can add counter-arguments to the graph. With the
examples above and with no method declarations a
\passthrough{\lstinline!*Thought!} belonging to one user cannot be
modified by a third party even though the fields of a
\passthrough{\lstinline!Thought!} struct is exposed due to Gno (runtime)
interrealm rules that taint third party reads via direct dot-selectors
\& index-expressions with a readonly-taint that persists even with
(direct selector) access of sub-fields.

The \passthrough{\lstinline!*Thought!} object can however be modified by
another user by calling a declared method. We can extend the
\passthrough{\lstinline!Thought!} struct with additional fields for
authorization or ownership and implement a method such as follows:

\begin{lstlisting}[language=Go]
type Thought struct {
    Owner        account
    Statement    string
    TypedSupport []*Thought
}

func (th *Thought) AddCounterArgument(cth *Thought) {
    caller := runtime.CurrentRealm().Address
    if th.Owner != caller {
        panic("unauthorized")
    }
    th.TypedSupport = append(th.TypedSupport,
        TypedThought{Type: "counter", Thought: cth})
}
\end{lstlisting}

This works but not well--it only works if the owner of the parent node
wants the counter-argument to be registered. Even if counter-arguments
were not registered as an assocation on chain, it is still possible for
any Gno.land state indexer to separately index the reverse association
of reference to the original \passthrough{\lstinline!*Thought!} when it
finds a counter-argument \passthrough{\lstinline!*Thought!} that
references in its struct field the original as a counter-argument. This
reliance on an external indexer shifts trust from the blockchain itself
to the indexer so is not always ideal.

Gno is intended for permissionless iteration and improvement so we will
discuss another way (among many) to manage associations of competing
thought statements; the pair-wise association among competing thought
statements can be registered in another (neutral) external realm that
allows the registration only at least one of the two thought statements
identify the other as a counter-argument. In this case it is not
necessary for a \passthrough{\lstinline!*Thought!} object to be
associated with any owner explicitly (via the
\passthrough{\lstinline!.Owner!} field). Note however that given the Gno
inter-realm specification to make a \passthrough{\lstinline!*Thought!}
object truly immutable even for the owner of the realm in which it
resides it must not expose any mutator functions, or it should have at
least a \passthrough{\lstinline!readonly bool!} field.

We can also add discussion board objects for each thought statement.

\begin{lstlisting}[language=Go]
// Board defines a type for boards.
type Board struct {
    // ID is the unique identifier of the board.
    ID ID
    // Name is the current name of the board.
    Name string
    // Aliases contains a list of alternative names for the board.
    Aliases []string
    // Readonly indicates that the board is readonly.
    Readonly bool
    // Threads contains all board threads.
    Threads PostStorage
    // ThreadsSequence generates sequential ID for new threads.
    ThreadsSequence IdentifierGenerator
    // Permissions enables support for permissioned boards.
    // This type of boards allows managing members with roles and permissions.
    // It also enables the implementation of permissioned execution of board related features.
    Permissions Permissions
    // Creator is the account address that created the board.
    Creator address
    // Meta allows storing board metadata.
    Meta any
    // CreatedAt is the board's creation time.
    CreatedAt time.Time
    // UpdatedAt is the board's update time.
    UpdatedAt time.Time
}

// New creates a new basic non permissioned board.
func New(id ID) *Board {
    return &Board{
        ID:              id,
        Threads:         NewPostStorage(),
        ThreadsSequence: NewIdentifierGenerator(),
        CreatedAt:       time.Now(),
    }
}
\end{lstlisting}

While it is certainly possible to embed a
\passthrough{\lstinline!*Board!} as a field of each
\passthrough{\lstinline!*Thought!}, the current implementation of
\passthrough{\lstinline!*Board!} is only safe from a moderation
perspective when it is permissioned; and so a board tightly coupled to a
\passthrough{\lstinline!*Thought!} may not be ideal depending on the
use-case. Instead we can map an external realm persisted index of
\passthrough{\lstinline!*Thought!} to \passthrough{\lstinline!*Board!}
associations similarly to how counter-thoughts are associated as
mentioned before. In both cases we probably want to add to the
\passthrough{\lstinline!Thought!} struct a globally unique ID like how
\passthrough{\lstinline!Board!} has. \emph{In the future we may provide
a standard function to get a unique identifier for every pointer object
but this has not yet been decided yet.}

Finally, consider for example the numbered sequence of verses of a book
of the bible, or the deep tree of statements in Wittgenstein's
Tractatus. In order to facilitate the efficient forking of such large
lists or graphs of objects it is necessary to avoid the usage of slices.
Even the avl.Tree (as provided in the Gno monorepo under the examples
directory) is not sufficient as it is a mutable tree. However a fork of
the avl.Tree into an immutable tree (or likewise a port of the iavl tree
in the tm2 Tendermint2 directory) can be used with some improvement to
allow for splicing in new elements and deleting existing elements from
the original tree.

So far I have illustrated a way for multiple users to construct their
thought statement graphs independently while also allowing for
counter-arguments to be registered/associated permissionlessly. More
design and exploration is needed to create a fully functional
permissionlessly extensible thought statement graph system; and in the
primordial soup of Gno ecosystem eventually one or more designs will
become dominant in usage by evolution. The reader is encouraged to
explore the above template and measure success by references and by
forks. See also \hyperref[use-case-95-facts]{Use Case: 95 Facts}.

\subsection{Use Case: Home Computing
Chains}\label{use-case-home-computing-chains}

The Gno VM is not just useful in the context of public decentralized
blockchains. It is also useful for home computing. Take for example
Email which despite all attempts to replace it still persists in our
lives today as flawed, complex, and outdated as it is. While a realm
that stores mail on the blockchain is not useful unless the data is
encrypted \emph{(and even if it were encrypted it is not a good idea nor
encouraged to store encrypted data on gno.land as encryption keys may
eventually get acquired by hackers and leaked and even persisted on the
blockchain too)}, the Gno VM can run anywhere, even on your private
server hosted at home. In fact, this is what we should do given the
prevalence of surveillance technology such as Google's Gmail which uses
AI to sort your mail and analyze for targeted advertising.

Imagine a black box local GnoVM you run at home. You can have the
\passthrough{\lstinline!/r/home/email!} realm store your emails at home
on your own home server. The same blockchain node logic can run on its
own as a single-validator \emph{home chain} which naturally supports
backups as secondary full nodes, or you can even make your home chain
byzantine fault-tolerant for better uptime.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Install in your home GnoVM chain a service plugin: /s/email/indexer,
  not /p/* nor /r/* but /s/* for off-chain service applications.
  \emph{(this prefix is not supported in gno.land but may be in the
  future)}.
\end{enumerate}

\begin{itemize}
\item
  /s/email/indexer reads state upon init, but also registers as a
  listener for notifications from /r/emails.
\item
  When a new email comes in, /r/emails via listeners calls
  \passthrough{\lstinline!/s/email/indexer.AddEmail()!}.
\item
  /s/email/indexer also imports /d/email/indexer which is an off-chain
  daemon component. Here /d/* represents a hypothetical prefix for Gno
  code to be run off-chain with arbitrary Go native functions available
  for import that would otherwise not be possible on gno.land (since a
  blockchain can only support deterministic logic).
\item
  /d/email/indexer can only access /s/email/indexer by a Gno firewall
  system declared with Gno package paths, types, and function/method
  names.
\item
  /s/email/indexer can import any /r/* or /p/* but not any /s/* (like
  Chrome extensions) and its own /d/email/indexer, unless otherwise
  restricted by the Gno firewall system.
\item
  Your mobile device registers an account with your local GnoVM home
  chain. This phone account is restricted to only access
  /s/email/indexer.
\item
  Your phone makes a request to /s/email/indexer. It then asks
  /d/email/indexer which in turn queries the local index and responds
  via /s/email/indexer.
\end{itemize}

Here are some benefits of GnoVM home computing: * Gno.land can be
leveraged to ensure that all software is properly audited. * Software is
expected to become finished and immutable. * All software benefits from
Go/Gno's type-safety and memory-safety. * A unified IPC system
drastically reduces surface area for penetration. * Plugin services and
daemons such as the aforementioned email indexer can be containerized
and restricted from unauthorized access. * Fine-grained
security/firewall rules at the function invocation level. * Byzantine
fault-tolerance comes out of the box for zero downtime. * Inversion of
control with public key cryptography for everything: no more password
management. * Resilience against infrastructure failures: home chains
continue operating during internet outages, power grid disruptions, or
state-level network shutdowns. Combined with mesh networking and local
power (solar, battery), GnoVM home chains can form the backbone of
community-level communication and coordination when centralized
infrastructure fails.

\subsection{Use Case: Supply Chain Transparency and Food
Sovereignty}\label{use-case-supply-chain-transparency-and-food-sovereignty}

The fragility of global supply chains has become impossible to ignore.
From semiconductor shortages to food supply disruptions, dependency on
opaque, centralized, and geographically concentrated supply chains puts
populations at the mercy of forces they cannot see, verify, or
influence. The geopolitical tension surrounding Taiwan and TSMC, the
concentration of pharmaceutical manufacturing in China and India, and
the consolidation of food production under a handful of multinational
corporations all represent single points of failure with civilizational
consequences.

Gno contracts can implement transparent, accountable supply chain
tracking where every step from origin to consumer is recorded with
cryptographic accountability. Unlike centralized supply chain solutions
(which are only as trustworthy as the company operating them), Gno-based
supply chain modules are open source, auditable, and composable. A food
provenance realm can import a certification realm which imports a
reputation realm--all with type-safe interoperability and no middleware.

For food sovereignty specifically, local communities can use Gno.land to
coordinate direct farmer-to-consumer networks, manage
community-supported agriculture (CSA) memberships, and maintain
transparent records of growing practices--all without depending on any
centralized platform that could deplatform them or change terms of
service. When combined with home computing chains, local food networks
can operate entirely independently of the global internet if necessary.

\subsection{Other Use Cases}\label{other-use-cases}

Gno.land can be used to host any other smart-contract application
supported by Ethereum written in Solidity, such as Defi applications,
name-resolution systems, DAOs and governance applications, etc.

You can explore the various dapps including sample implementations of
ERC equivalents in the
\href{https://github.com/gnolang/gno/tree/master/examples/gno.land}{examples
directory}. \emph{Note that these prototypes have not yet been audited
unless otherwise specified!}

There are also many good use-cases for integrating the GnoVM in client
or server side programming where deterministic gas-limited (for CPU,
disk, memory) execution of logic is desired. In particular Gno and the
Gno VM can be ideal for scripting in the gaming context for multi-player
games. There are many games like Minecraft (with JavaScript) or Second
Life (with LSL) that offer scripting in the game, but they are usually
not both deterministic and gas-counted; but these are the properties you
may need in a multi-player game if you also want to provide
fault-tolerant, accountability, and replayability guarantees (with or
without decentralization or replication).

The main feature of Gno is its multi-user context awareness. Where-ever
you want user provided logic directly interacting with those of another
user Gno solutions will be more succinct, elegant, and easy to develop
and maintain.

\subsection{Gno.land Constitution}\label{gno.land-constitution}

See {[}./CONSTITUTION.md{]} for the Gno.land Constitution (draft) and
details of genesis, tokenomics, governance, and more.

\paragraph{Separation of Church and
State}\label{separation-of-church-and-state}

Madison separated church and state in the US Constitution albeit there
is a hint of the Christian spirit by the way in which the constitution
was signed: ``\ldots{} in the Year of the Lord\ldots{}''. All the
founders were Christian including Jefferson and Madison, and in
particular the primary author of the US Constitution James Madison
explicitly separated church from the constitution so as to help promote
the teachings of Jesus as evidenced in his other writings. Likewise
Gno.land besides this whitepaper is independent of any religion by its
constitution, which should only refer to this whitepaper sparingly.

Gno.land will launch with a minimal (living) constitution written and
maintained in English, but also ultimately be supplemented by the
completed GnoVM code and Tendermint2 and Gno.land implementation. Future
implementations of the GnoVM and Gno.land should adhere to the completed
software mentioned above.

Gno.land should not censor speech, even if the speech is wrong. However,
it should ban all porn and try to limit external links to porn sites as
porn is not speech and is dangerous to civilization. Whether hate-speech
is tolerated should be determined by each realm but also by the living
Gno.land constitution and by GovDAO vote to amend the constitution and
laws of Gno.land.

\subsection{GnoWeb Browser}\label{gnoweb-browser}

GnoWeb is the server software for Gno.land, a browser within a browser
for viewing realm data.

Instead of requiring realm applications to return HTML, the convention
is to implement a Render() function that returns Markdown. This is to
allow the transition away from the bloated HTML standards and browser
software and realm data to be browsed even from the console. Note that
we don't need HTML XML elements to denote objects: everything in Gno is
already an object. This makes Gno.land more like the original World Wide
Web that conforms to the \emph{Document Object} model (DOM).

GnoWeb does support some custom XML elements for improved layout and
functionality, such as column layout and form submissions that integrate
with browser extensions for transaction signing.

Note that GnoWeb is not yet a general blockchain explorer (e.g.~for
transactions) nor a general purpose Gno.land state explorer. A Gno.land
specific blockchain explorer already exists. GnoWeb can only render
markdown returned from Render() functions, and a general purpose state
explorer is still desired.

Realm code is not precluded from returning HTML or even JSON for custom
browser applications. In the near future the Gno.land node software will
support returning JSON encodings of Gno objects. Thus future alternative
browser applications may provide more interactive rich user experiences
for viewing and mutating Gno.land state without any Markdown
intermediary representation; and perhaps leveraging AI for intelligent
layout and styling for rich interactivity.

\newpage

\section{Future Work}\label{future-work}

\begin{itemize}
\item
  \textbf{Access control systems.} A composable library of membership,
  reputation, and moderation primitives that realms can import and
  combine without any single mandated identity system. Components
  include web-of-trust scoring, deposit-based access, viral invitation
  trees, hierarchical accountability chains, and oracle-backed
  attestations. The goal is a free market of access control policies
  rather than a universal gatekeeping layer.
  \href{https://github.com/jaekwon/gnoauth}{GnoAuth} is a first step: an
  OAuth identity oracle that binds Web2.0 accounts (e.g.~GitHub) to Gno
  addresses with signed attestations and sybil-resistance metadata.
\item
  \textbf{Persistent garbage collector.} The current implementation
  persists objects with reference counting and deletes zero-ref-count
  objects at realm-transaction finalization, but does not yet collect
  reference cycles. A full persistent cycle-detecting garbage collector
  will reclaim storage for cyclic structures that become unreachable. A
  transaction-scoped cyclic GC may also be implemented as a lighter
  alternative for common cases.
\item
  \textbf{Name registry for realm and package paths.} A decentralized
  naming system for mapping human-readable names to realm and package
  paths, enabling stable references even when packages are re-deployed
  or versioned. This would function similarly to DNS but for the Gno
  package namespace.
\item
  \textbf{Realm upgrading.} A mechanism for upgrading mutable realm code
  while preserving existing state and object references. This requires
  careful design to maintain type safety across upgrades and to ensure
  that dependent realms are not broken by upstream changes. Migration
  functions and versioned type schemas are under consideration.
\item
  \textbf{Realm data browser.} A general-purpose state explorer for
  inspecting any realm's persisted object graph, not limited to Render()
  output. This would allow developers and users to browse the full
  object tree, inspect individual objects, view ownership and reference
  graphs, and trace cross-realm references.
\item
  \textbf{Deterministic concurrency.} Extending Gno with deterministic
  concurrency primitives that preserve the reproducibility required for
  consensus while enabling parallel execution of independent realm
  transactions. This could unlock significant throughput improvements
  without sacrificing the determinism guarantee.
\item
  \textbf{Joeson parser.} A general-purpose parser library implemented
  in Gno for on-chain parsing of structured data formats,
  domain-specific languages, and protocol messages. This enables realms
  to define and interpret custom grammars without relying on off-chain
  tooling.
\item
  \textbf{Gno2.} A future major revision of the Gno language
  incorporating lessons learned from production use. Potential additions
  include a \passthrough{\lstinline!readonly!} modifier for function
  return types, native syntax for \passthrough{\lstinline!cross!} and
  \passthrough{\lstinline!cur realm!} (replacing the current overloaded
  function argument convention), and further language-level support for
  multi-user patterns that emerge from ecosystem usage.
\item
  \textbf{Open hardware.} Memristor-based memory systems eliminate the
  distinction between volatile RAM and persistent storage, which aligns
  naturally with the GnoVM's automatic persistence model. Purpose-built
  hardware could further simplify the VM and dramatically improve
  performance. Open hardware designs for validator nodes would also
  strengthen decentralization by reducing dependence on proprietary chip
  manufacturers.
\end{itemize}

\end{document}
